---
title: "Multi-state analysis and creation of the MSMplus input files with Stata"
output: html_document
---


```{r setup,results = 'hide',message = FALSE, warning = FALSE}
library("MSMplus")
library("survival")
library("mstate")
library("dplyr")
library("reshape2")
library("flexsurv")
library("msm")
library("RJSONIO")
library("shiny")
library("knitr")


```


The advisable approach is the manual creation of an excel/csv file with the analysis results by the researcher according to certain formatting and naming rules described in this [tutorial](https://nskbiostatistics.shinyapps.io/supplementary/). We have created an alternative way for deriving the MSMplus input files to avoid the labour. The json files can be easily derived while running the multi-state models. In Stata, this is done via the commands msboxes and predictms and in R via the current package and the use of its funtions: flexjson, msmjson, mstatejson. That being said we still advise for the manual approach as the researcher does not need to have any knowledge of R or Stata, as the analysis can be conducted via any statistical software.


The user can locally launch the MSMplus application by writting MSMplus_prepare::runMSMplus() or access it online at https://nskbiostatistics.shinyapps.io/MSMplus/


# Files input preparation

## Function msboxes_R
### Generating information to create a multi-state graph with updated frequencies in each state across different time points

We will start by using data from the European Blood and Marrow Transplant registry. The dataset consists of 2204 patients who received bone marrow transplantation. The three states a patient can be in is 1) Post- transplant, 2) Platelet recovery 3) Relapse/Death. The covariate patterns used in this example are the 3 age categories, namely <20 y.old, 20-40y.old and >40 y.old . This dataset is freely available from mstate package and you can access more information by typing ?ebmt3.

```{r, echo = TRUE}
load("ebmt.rda", envir = parent.frame(), verbose = FALSE)

head(ebmt)
```



 Let's first define the transition matrix
```{r, echo = TRUE}
tmat <- transMat(x = list(c(2, 3),c(3), c() ), names = c("Transplant", "Platelet Recovery", "Relapse/Death" ) )
```


We will now create dummy variables for the age categories
```{r, echo = TRUE}
ebmt$age2=  recode(ebmt$age, ">40" =0, "20-40"=1,"<=20" =0 )
ebmt$age3=  recode(ebmt$age, ">40" =1, "20-40"=0,"<=20" =0 )
```

Data preparation- From one row per participant to multiple rows per participant, one for each allowed transition.
```{r, echo = TRUE}
msebmt <- msprep(data = ebmt, trans = tmat, 
                 time = c(NA, "prtime", "rfstime"), status = c(NA, "prstat", "rfsstat"), keep=c("age2","age3"))

head(msebmt)
```

We can now call function msboxes_R
 
msboxes_R will create a json file containing parameters that will help MSMplus to automatically create 
the multi-state graph of each specific setting. However, the user has the option to design and create the
multistate graph within the app as well. 

```{r, echo = TRUE,message = FALSE }

results3_days=MSMplus::msboxes_R(data=msebmt,id= msebmt$id, yb=c(0.3,0.5,0.75),
                        xb=c(0.5,0.2,0.7),boxwidth=0.1,boxheight=0.1,
                        tmat.= tmat, tstop=msebmt$Tstop,vartime=c(seq(0,10,by=1)),scale=365.25,
                        jsonpath="~", name="msboxes_EBMT.json" ) 
results3_days

```




## MSMplus Json input file of predictions: The flexjson function 

 Provide time vector
```{r, echo = TRUE}
tgrid <- seq(1, 10, by = 1)   
```
Provide transition matrix
```{r, echo = TRUE}
tmat <- rbind(c(NA, 1, 2), c(NA, NA, 3), c(NA, NA, NA)) 
```

 Run transition specific hazard models: Clock forward approach and use of flexible parametric models
```{r, echo = TRUE}

cfwei.list<-vector(3,mode="list")

for (i in 1:3) {
  
  cfwei.list[[i]]<-flexsurvreg(Surv(Tstart,Tstop,status)~age2+age3,subset=(trans==i),
                               dist="weibull",data=msebmt)
}

  
```


 Prediction for different covariate patterns (the 3 age categories)
```{r, echo = TRUE}
wh1 <- which(msebmt$age2 == 0 & msebmt$age3 == 0)
pat1 <- msebmt[rep(wh1[1], 3), 9:10]
attr(pat1, "trans") <- tmat


wh2 <- which(msebmt$age2 == 1 & msebmt$age3 == 0)
pat2 <- msebmt[rep(wh2[1], 3), 9:10]
attr(pat2, "trans") <- tmat

wh3 <- which(msebmt$age2 == 0 & msebmt$age3 == 1)
pat3 <- msebmt[rep(wh3[1], 3), 9:10]
attr(pat3, "trans") <- tmat
```

We now run the flexsurvjson function to perform the multi-state model analysis using the function 
from package flexsurv and the pack the predictions in a json file.
```{r, echo = TRUE}
results_cf <- MSMplus::flexsurvjson( model=cfwei.list, vartime=seq(365.25,365.25,by=365.25), 
                                   qmat=tmat, process="Markov",
                                   totlos=TRUE, ci.json=FALSE, cl.json=0.95, B.json=10, tcovs=NULL,
                                   Mjson=100, variance=FALSE,
                                   covariates_list=list(pat1,pat2,pat3), 
                                   jsonpath="~",
                                   name="predictions_EBMT_flex.json" ) 
```


```{r, echo = TRUE}
results_cf$timevar
results_cf$Nats
results_cf$atlist
results_cf$tmat
results_cf$Ntransitions
results_cf$is.cumhaz
results_cf[[7]]
```

If the user has used the clock reset approach they have to specify "semiMarkov" at the process argument.
